---
title: "210531_Analysis_of_FNANA"
output: html_notebook
---

```{r}
#Install packages:
install.packages("ggfortify")
install.packages("calibrate")
install.packages("BiocManager")
BiocManager::install(c("edgeR"))
BiocManager::install(c("biomaRt"))
BiocManager::install(c("DOSE"))
BiocManager::install(c("clusterProfiler","Rgraphviz","DOSE"))
BiocManager::install(c("org.Mm.eg.db"))
```

#### re-install originally used versions for GO analysis
```{r}
#used IRanges_2.18.3
packageurl <- "http://bioconductor.riken.jp/packages/3.9/bioc/bin/macosx/el-capitan/contrib/3.6/IRanges_2.18.3.tgz"
install.packages(packageurl, repos=NULL, type="source")
```

```{r}
#used S4Vectors_0.22.1
packageurl <- "http://bioconductor.riken.jp/packages/3.9/bioc/bin/macosx/el-capitan/contrib/3.6/S4Vectors_0.22.1.tgz"
install.packages(packageurl, repos=NULL, type="source")
```

```{r}
#used BiocGenerics_0.30.0
packageurl <- "http://bioconductor.riken.jp/packages/3.9/bioc/src/contrib/BiocGenerics_0.30.0.tar.gz"
install.packages(packageurl, repos=NULL, type="source")
```

```{r}
#Install old version of org.Mm.eg.db (used 3.8.2).
packageurl <- "http://bioconductor.riken.jp/packages/3.9/data/annotation/src/contrib/org.Mm.eg.db_3.8.2.tar.gz"
install.packages(packageurl, repos=NULL, type="source")
```

```{r}
#used clusterProfiler_3.12.0 on 190828
packageurl <- "http://bioconductor.riken.jp/packages/3.9/bioc/src/contrib/clusterProfiler_3.12.0.tar.gz"
install.packages(packageurl, repos=NULL, type="source")
```

```{r}
#used DOSE_3.10.2
packageurl <- "http://bioconductor.riken.jp/packages/3.9/bioc/src/contrib/DOSE_3.10.2.tar.gz"
install.packages(packageurl, repos=NULL, type="source")
```

```{r}
#used AnnotationDbi_1.46.1
packageurl <- "http://bioconductor.riken.jp/packages/3.9/bioc/src/contrib/AnnotationDbi_1.46.1.tar.gz"
install.packages(packageurl, repos=NULL, type="source")
```

```{r}
#used Biobase_2.44.0
packageurl <- "http://bioconductor.riken.jp/packages/3.9/bioc/bin/macosx/el-capitan/contrib/3.6/Biobase_2.44.0.tgz"
install.packages(packageurl, repos=NULL, type="source")
```

```{r}
#used biomaRt_2.40.5
packageurl <- "http://bioconductor.riken.jp/packages/3.9/bioc/src/contrib/biomaRt_2.40.5.tar.gz"
install.packages(packageurl, repos=NULL, type="source")
```

#### Load packages:
```{r}
library(edgeR)
library(ggplot2)
library(ggfortify)
library(biomaRt)
library(calibrate)
library(DOSE)
library(clusterProfiler)
library(org.Mm.eg.db)
library(tidyr)
library(dplyr)
library(ggrepel)
library(pheatmap)
```

```{r}
sessionInfo()
```

#### Load data:
```{r}
counts <- read.delim("~/Desktop/171106_FNANA.clean.readCount", row.names = 1)
head(counts)
```

#### Remove genes without any counts in any samples:
```{r}
counts_wco=counts[(rowSums(counts>0)>=1),]
```

#### Select 0h columns:
```{r}
cols_0h=c(1,3,6,8,10,13,15,17,20)
hour0_SS=counts_wco[,cols_0h]
head(hour0_SS)

#remove additional zero counts:
hour0_SS_wco=hour0_SS[(rowSums(hour0_SS>0)>=1),]
```

#### edgeR (DEG) analysis
```{r}
#For easy manipulation, we put the data into a DGEList object:
y_0h=DGEList(hour0_SS_wco)

#TMM normalization is applied to this dataset to account for compositional difference between the libraries.
y_0h = calcNormFactors(y_0h)
```

```{r}
#Before we fit negative binomial generalized linear models (GLMs), we need to define our design matrix based on the experimental design:
Experiment <- factor(c(rep(c(1,2,3), each=3)))

Treatment <- factor(c("F.NANA","GM.CSF","P.NANA","F.NANA","GM.CSF","P.NANA","F.NANA","GM.CSF","P.NANA"))

design_0h <- model.matrix(~Experiment+Treatment)

rownames(design_0h) <- colnames(y_0h)

#print the design matrix:
design_0h
```

```{r}
#First we estimate the overall dispersion for the dataset, to get an idea of the overall level of biological variability:
y_0h <- estimateGLMCommonDisp(y_0h, design=design_0h, verbose=TRUE)

#Now proceed to determine differentially expressed genes. Fit genewise GLMs
fit_0h <- glmFit(y_0h, design_0h)
```

```{r}
#F.NANA is base!

#Conduct likelihood ratio tests for differences in treatments and show the top genes:
lrt_0h_GvsF <- glmLRT(fit_0h, coef=4)
topTags(lrt_0h_GvsF)
summary(decideTestsDGE(lrt_0h_GvsF))
```

#### Add official gene symbols to lists and write the results of the DE analysis to a table:
```{r}
results_0h_GvsF <- topTags(lrt_0h_GvsF,n = length(y_0h$counts[,1]))
```

```{r}
mart <- useMart("ENSEMBL_MART_ENSEMBL", dataset = "mmusculus_gene_ensembl",host="www.ensembl.org")
ensembl2name <- getBM(attributes=c("ensembl_gene_id","external_gene_name"),mart=mart)
```

```{r}
results_0h_GvsF <- merge(x=results_0h_GvsF$table, y=ensembl2name, by.x=0, by.y=1, all.x=TRUE)
```

```{r}
write.table(as.matrix(results_0h_GvsF),file="~/Desktop/200120_0h_GvsF_DEG",sep="\t")
```

#### Export normalized count values
```{r}
#write.table(as.matrix(fit_0h$fitted.values),file="~/Desktop/200120_0h_norm.counts",sep="\t")
```

#### Reload the results
```{r}
results_0h_GvsF = read.delim("~/Desktop/200120_DEG_and_normCounts/200120_0h_GvsF_DEG",sep="\t")
```

```{r}
#select significant genes by FDR:
results_0h_GvsF_FDRsig = results_0h_PvsF[results_0h_GvsF$FDR<0.05,]
nrow(results_0h_GvsF_FDRsig)
```

```{r}
#select top 250 up/down genes:
results_0h_GvsF_FDRsig_top250 = bind_rows(top_n(results_0h_GvsF_FDRsig, 250, logFC), top_n(results_0h_GvsF_FDRsig, -250, logFC))
```


#Gene Ontology on top 250 up or downregulated significant genes
###Note: It's better to use ENSEMBL or ENTREZ IDs for the GO analysis.

#### Biological Process GO analysis
```{r}
results_0h_GvsF_FDRsig_top250_BP <- enrichGO(gene = results_0h_GvsF_FDRsig_top250$Row.names,
                keyType       = 'ENSEMBL',
                OrgDb         = org.Mm.eg.db,
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.05,
                minGSSize     = 3,
                readable      = TRUE)
head(summary(results_0h_GvsF_FDRsig_top250_BP))
```

```{r}
##write the analysis results to file:
write.table(as.matrix(summary(results_0h_GvsF_FDRsig_top250_BP)),file="~/Desktop/200507_results_0h_GvsF_FDRsig_top250_BP.txt",sep="\t")
```


#### Molecular Functions GO analysis
```{r}
results_0h_GvsF_FDRsig_top250_MF <- enrichGO(gene = results_0h_GvsF_FDRsig_top250$Row.names,
                keyType       = 'ENSEMBL',
                OrgDb         = org.Mm.eg.db,
                ont           = "MF",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.05,
                minGSSize     = 3,
                readable      = TRUE)
head(summary(results_0h_GvsF_FDRsig_top250_MF))
```

```{r}
#write the analysis results to file:
write.table(as.matrix(summary(results_0h_GvsF_FDRsig_top250_MF)),file="~/Desktop/200201_results_0h_GvsF_FDRsig_top250_MF.txt",sep="\t")
```

#### re-load results
```{r}
#reload GO/KEGG results
results_0h_GvsF_FDRsig_top250_BP = read.delim("~/Desktop/NANA_input/200201_GO_KEGG_FDR_Top250_Up_Down/200201_results_0h_GvsF_FDRsig_top250_BP.txt",sep="\t", row.names = 1)
results_0h_GvsF_FDRsig_top250_MF = read.delim("~/Desktop/NANA_input/200201_GO_KEGG_FDR_Top250_Up_Down/200201_results_0h_GvsF_FDRsig_top250_MF.txt",sep="\t", row.names = 1)
```

```{r}
#add column for decimal number of GeneRatio
results_0h_GvsF_FDRsig_top250_MF = separate(data = results_0h_GvsF_FDRsig_top250_MF, col = GeneRatio, into = c("left", "right"), sep = "/")
results_0h_GvsF_FDRsig_top250_MF$GeneRatio_decimal = (
   as.numeric(results_0h_GvsF_FDRsig_top250_MF$left)/as.numeric(results_0h_GvsF_FDRsig_top250_MF$right))

results_0h_GvsF_FDRsig_top250_BP = separate(data = results_0h_GvsF_FDRsig_top250_BP, col = GeneRatio, into = c("left", "right"), sep = "/")
results_0h_GvsF_FDRsig_top250_BP$GeneRatio_decimal = (
   as.numeric(results_0h_GvsF_FDRsig_top250_BP$left)/as.numeric(results_0h_GvsF_FDRsig_top250_BP$right))
```


```{r}
y <- as.data.frame(results_0h_GvsF_FDRsig_top250_MF)
head(y)
y$Description <- factor(y$Description, levels = y$Description[order(y$GeneRatio_decimal)])
y = y[0:15,]
y = na.omit(y)
gg1 = ggplot(y,
      aes(x = 1, y = Description)) + 
      geom_point(aes(size = Count, color = p.adjust))+# color = "black", pch = 21) +
      theme_bw(base_size = 14) +
      scale_colour_gradient(limits = c(0,0.05), low = "darkgreen", high="white", guide = guide_colorbar(frame.colour = "black", ticks.colour = "black")) +
      geom_point(aes(size = Count), color = "black", pch = 21) +
      ylab(NULL) +
      xlab(NULL) +
      scale_x_discrete(expand=c(0,0))  +
      guides(size = guide_legend(order = 1)) +
      labs(size = "genes", color = "p value") +
      theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
      ggtitle("0h_GvsF_FDRsig_top250_MF") +
      theme(plot.title = element_text(hjust=1))
gg1
```

```{r}
pdf("~/Desktop/210102_0h_GvsF_FDRsig_top250_MF.pdf",width=6.2,height=3.6,paper='special') 
gg1
dev.off()
```

```{r}
y <- as.data.frame(results_0h_GvsF_FDRsig_top250_BP)
head(y)
y$Description <- factor(y$Description, levels = y$Description[order(y$GeneRatio_decimal)])
y = y[0:15,]
y = na.omit(y)
gg1 = ggplot(y,
      aes(x = 1, y = Description)) + 
      geom_point(aes(size = Count, color = p.adjust))+# color = "black", pch = 21) +
      theme_bw(base_size = 14) +
      scale_colour_gradient(limits = c(0,0.05), low = "darkgreen", high="white", guide = guide_colorbar(frame.colour = "black", ticks.colour = "black")) +
      geom_point(aes(size = Count), color = "black", pch = 21) +
      ylab(NULL) +
      xlab(NULL) +
      scale_x_discrete(expand=c(0,0))  +
      guides(size = guide_legend(order = 1)) +
      labs(size = "genes", color = "p value") +
      theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
      ggtitle("0h_GvsF_FDRsig_top250_BP") +
      theme(plot.title = element_text(hjust=1))
gg1
```

```{r}
pdf("~/Desktop/210102_0h_GvsF_FDRsig_top250_BP.pdf",width=6.2,height=3.6,paper='special') 
gg1
dev.off()
```

### simplify GO BP results
```{r}
#re-load results
results_0h_GvsF = read.delim("~/Desktop/200120_DEG_and_normCounts/200120_0h_GvsF_DEG",sep="\t")
```

```{r}
#select significant genes by FDR:
results_0h_GvsF_FDRsig = results_0h_PvsF[results_0h_GvsF$FDR<0.05,]
nrow(results_0h_GvsF_FDRsig)
```

```{r}
#select top 250 up/down genes:
results_0h_GvsF_FDRsig_top250 = bind_rows(top_n(results_0h_GvsF_FDRsig, 250, logFC), top_n(results_0h_GvsF_FDRsig, -250, logFC))
```

```{r}
#re-run the GO analysis
results_0h_GvsF_FDRsig_top250_BP <- enrichGO(gene = results_0h_GvsF_FDRsig_top250$Row.names,
                keyType       = 'ENSEMBL',
                OrgDb         = org.Mm.eg.db,
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.05,
                qvalueCutoff  = 0.05,
                minGSSize     = 3,
                readable      = TRUE)
head(summary(results_0h_GvsF_FDRsig_top250_BP))
```


```{r}
#simplify results
results_0h_GvsF_FDRsig_top250_BP_simple30 = simplify(
  results_0h_GvsF_FDRsig_top250_BP,
  cutoff = 0.3,
  by = "p.adjust",
  select_fun = min,
  measure = "Wang",
  semData = NULL
)
```

```{r}
#export and re-import results
write.table(as.matrix(summary(results_0h_GvsF_FDRsig_top250_BP_simple30)),file="~/Desktop/200507_results_0h_GvsF_FDRsig_top250_BP_simple30.txt",sep="\t")
```

```{r}
results_0h_GvsF_FDRsig_top250_BP_simple30 = read.delim("~/Desktop/200507_results_0h_GvsF_FDRsig_top250_BP_simple30.txt",sep="\t", row.names = 1)
```


```{r}
#add column for decimal number of GeneRatio
results_0h_GvsF_FDRsig_top250_BP_simple30 = separate(data = results_0h_GvsF_FDRsig_top250_BP_simple30, col = GeneRatio, into = c("left", "right"), sep = "/")
results_0h_GvsF_FDRsig_top250_BP_simple30$GeneRatio_decimal = (
   as.numeric(results_0h_GvsF_FDRsig_top250_BP_simple30$left)/as.numeric(results_0h_GvsF_FDRsig_top250_BP_simple30$right))
```

```{r}
y <- as.data.frame(results_0h_GvsF_FDRsig_top250_BP_simple30)
head(y)
y$Description <- factor(y$Description, levels = y$Description[order(y$GeneRatio_decimal)])
y = y[0:15,]
y = na.omit(y)
gg2 = ggplot(y,
      aes(x = 1, y = Description)) + 
      geom_point(aes(size = Count, color = p.adjust))+# color = "black", pch = 21) +
      theme_bw(base_size = 14) +
      scale_colour_gradient(limits = c(0,0.05), low = "darkgreen", high="white", guide = guide_colorbar(frame.colour = "black", ticks.colour = "black")) +
      geom_point(aes(size = Count), color = "black", pch = 21) +
      ylab(NULL) +
      xlab(NULL) +
      scale_x_discrete(expand=c(0,0))  +
      guides(size = guide_legend(order = 1)) +
      labs(size = "genes", color = "p value") +
      theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
      ggtitle("0h_GvsF_FDRsig_top250_BP_simple30") +
      theme(plot.title = element_text(hjust=1))
gg2
```

```{r}
pdf("~/Desktop/210507_0h_GvsF_FDRsig_top250_BP_simple30.pdf",width=6.2,height=3.6,paper='special') 
gg2
dev.off()
```


#### Beautify volcanoplots and highlight genes
```{r}
#Re-load data
results_0h_GvsF = read.delim("~/Desktop/NANA_input/200120_0h_GvsF_DEG",sep="\t", row.names = 1)
```

```{r}
# set values outside axis limits to Inf
upper_FDR_limit = 70
upper_logFC_limit = 15
lower_logFC_limit = -15
results_0h_GvsF_limits = results_0h_GvsF
results_0h_GvsF_limits$logFDR = -log10(results_0h_GvsF$FDR)
results_0h_GvsF_limits$logFDR[results_0h_GvsF_limits$logFDR > upper_FDR_limit] = Inf
results_0h_GvsF_limits$logFC[results_0h_GvsF_limits$logFC > upper_logFC_limit] = upper_logFC_limit
results_0h_GvsF_limits$logFC[results_0h_GvsF_limits$logFC < lower_logFC_limit] = lower_logFC_limit
```

```{r}
results_0h_GvsF = results_0h_GvsF_limits
```

```{r}
highlight_genes = c("kept secret until publication")
```

```{r}
results_0h_GvsF_sub = subset(results_0h_GvsF, abs(logFC)>log2(2) & FDR<0.05)
results_0h_GvsF_highlight = subset(results_0h_GvsF, results_0h_GvsF$external_gene_name %in% highlight_genes)

vp_sign_sub_up = subset(results_0h_GvsF, logFC>0 & FDR<0.05)
vp_sign_sub_dn = subset(results_0h_GvsF, logFC<0 & FDR<0.05)
vp_non_sign = subset(results_0h_GvsF, FDR>0.05 | abs(logFC)<1)

vp.results_0h_GvsF = ggplot(results_0h_GvsF) +
  geom_point(
      data = results_0h_GvsF,
      aes(x = logFC, y = logFDR),
      color = "grey",
      alpha = 0.2,
      cex = 1
    ) +
  geom_point(
      data = subset(vp_sign_sub_up, FDR<0.05),
      aes(x = logFC, y = logFDR),
      fill = "blue",
      alpha = 0.3,
      color = "darkblue",
      pch = 21
    ) +
  geom_point(
      data = subset(vp_sign_sub_dn, FDR<0.05),
      aes(x = logFC, y = logFDR),
      fill = "red",
      alpha = 0.3,
      color = "darkred",
      pch = 21
    ) +
    geom_text_repel(
      data = results_0h_GvsF_highlight,
      aes(x = logFC, y = logFDR, label=external_gene_name),
      size = 5,
      min.segment.length = 0, #use to always put a line,
      box.padding = unit(0.35, "lines"),
      point.padding = unit(0.3, "lines"),
      segment.size = 1
    ) +
    theme_bw(base_size = 18) +
    labs(x = "logFC", y = "-log10(FDR)") +
    guides(size=guide_legend(override.aes=list(fill="white"))) +
    ggtitle("vp.results_0h_GvsF")
vp.results_0h_GvsF
```

```{r}
pdf("~/Desktop/210531_0h_GvsF_Volcano.pdf",width=7,height=7,paper='special') 
vp.results_0h_GvsF
dev.off()
```


## Calculate TPM expression

####Load data:
```{r}
counts_wl <- read.delim("~/Desktop/171106_FNANA.clean.readCount.withLength", row.names = 1)
```

#### Remove genes without any counts in any samples:
```{r}
counts_wl_wco=counts_wl[(rowSums(counts_wl>0)>=1),]
```

#### Add official gene name to the table
```{r}
info <- read.delim("~/Desktop/gencode.mouse.v1.annotation.gene_info.txt", row.names = 1)
head(info)

counts_wl_wco_name = merge(x=counts_wl_wco, y=info, by.x=0, by.y=0, all.x=TRUE)
head(counts_wl_wco_name)
```

### TPM calculation
```{r}
counts_wl_tpm = counts_wl_wco_name[,3:23]
RPK = counts_wl_tpm/counts_wl$Length
RPK=RPK*1000
head(RPK)

sum_RPK = colSums(RPK)
sum_RPK

TPM= sweep(RPK,2,sum_RPK,`/`)
TPM=TPM*10^6
head(TPM)

TPM = cbind(counts_wl_wco_name[,c(1,25)],TPM)
head(TPM)

write.table(as.matrix(TPM),file="~/Desktop/210417_FNANA_TPM.txt",sep="\t")
```

```{r}
#reload TPM list
TPM <- read.delim("~/Desktop/210417_FNANA_TPM.txt", row.names = 1)
```


##Heatmaps of selected genes

####Load gene lists
```{r}
tlr_genes = read.delim("~/Desktop/210531_tlr_genes.txt", header = F)
tlr_genes$order = rownames(tlr_genes)
tlr_genes
tlr_genes$order = as.factor(c(01,02,03,04,05,06,07,08,09,10:13))
tlr_genes
```

#### Merge with TPM list
```{r}
heatmap_TPM_tlr = merge(x=tlr_genes, y=TPM, by.x=1, by.y=2, all.x=TRUE)
rownames(heatmap_TPM_tlr) = heatmap_TPM_tlr$V1

heatmap_TPM_tlr = as.data.frame(heatmap_TPM_tlr)
heatmap_TPM_tlr = heatmap_TPM_tlr[order(heatmap_TPM_tlr$order),]
```

#### select order of columns
```{r}
heatmap_TPM_tlr_order = cbind(heatmap_TPM_tlr$M1_GM.CSF_0h, heatmap_TPM_tlr$M2_GM.CSF_0h, heatmap_TPM_tlr$M3_GM.CSF_0h, heatmap_TPM_tlr$M1_F.NANA_0h, heatmap_TPM_tlr$M2_F.NANA_0h, heatmap_TPM_tlr$M3_F.NANA_0h)
rownames(heatmap_TPM_tlr_order) = rownames(heatmap_TPM_tlr)
colnames(heatmap_TPM_tlr_order) = c("M1_GM.CSF_0h", "M2_GM.CSF_0h", "M3_GM.CSF_0h", "M1_F.NANA_0h", "M2_F.NANA_0h", "M3_F.NANA_0h")
```

#### plotting
```{r, fig.width=3, fig.height=6}
breaksList = seq(0, 1, by = 0.001)

TPM_heat_tlr=pheatmap(heatmap_TPM_tlr_order, border_color = "black", scale = "row", cluster_rows = F, cluster_cols = F,
         color = colorRampPalette(c("white", "darkgreen"))(1000),
         breaks = breaksList)
```

```{r}
pdf("~/Desktop/210531_tlr_rowScaled.pdf",width=2.3,height=4,paper='special') 
TPM_heat_tlr
dev.off()
```

```{r, fig.width=3, fig.height=6}
breaksList = seq(0, 1200, by = 1)

TPM_heat_tlr=pheatmap(heatmap_TPM_tlr_order, border_color = "black", cluster_rows = F, cluster_cols = F,
         color = colorRampPalette(c("white", "darkgreen"))(1200),
         breaks = breaksList)
```

```{r}
pdf("~/Desktop/210520_tlr_rawTPM.pdf",width=2.3,height=4,paper='special') 
TPM_heat_tlr
dev.off()
```

```{r}
heatmap_TPM_tlr_order_log = log2(heatmap_TPM_tlr_order)
```

```{r, fig.width=3, fig.height=6}
breaksList = seq(0, 10, by = 0.001)

TPM_heat_tlr=pheatmap(heatmap_TPM_tlr_order_log, border_color = "black", cluster_rows = F, cluster_cols = F,
         color = colorRampPalette(c("white", "darkgreen"))(10000),
         breaks = breaksList)
```

```{r}
pdf("~/Desktop/210520_tlr_log2TPM.pdf",width=2.3,height=4,paper='special') 
TPM_heat_tlr
dev.off()
```

```{r}
sessionInfo()
```

